---
title: "Reproducible Research Course Project 1"
author: "Miguel Ángel García"
date: "2015/11/14"
output: html_document
---
# Introduction #
Using data from activity monitoring devices, the goal of this assignment is to answer different question about this data with the help of the literate statistical programming *knitr* package.

## The Dataset ##
We'll use the [Activity Monitoring Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) dataset, previously downloaded in the working directory.

### Structure ###
The first view of the data (using a simple text editor) shows a CSV file with a simple structure. Here are the first three rows:

> "steps","date","interval"  
> NA,"2012-10-01",0  
> NA,"2012-10-01",5  

As we can see, there are three columns in the dataset (the first row shows the names):

* **Steps:** Number of steps in the 5 minutes interval
* **Date:** Date of the measure (in YYYY-MM-DD format)
* **Interval:** 5 minutes interval of each measure

It's noticed that the *steps* column has NA values.

### Loading the Data ###
We'll use the *dplyr* package to transform and filter the data. And we'll load it by using the *read.csv* function with the argument *header = TRUE*, as we know that the data has column names in the first row.

```{r, echo= TRUE, warning= FALSE, results= 'hide'}
library(dplyr, warn.conflicts= FALSE, quietly= TRUE)
actDS <- tbl_df(read.csv("activity.csv", header= TRUE))
```

### Some Initial Checks ###
To verify that the data is properly loaded, we can do some simple validations. We'll start by printing the first 5 rows of the data:
```{r, echo= TRUE}
head(actDS, 5)
```

We can notice that the date column is not typed as *date*:
```{r, echo= TRUE}
typeof(actDS$date)
```

So we need to convert it:
```{r, echo= TRUE}
actDS$date <- as.Date(actDS$date)
```

# Questions #
Now, we have the dataset ready to analyze, so we start to answer the questions one by one.

## Number of Steps Taken per Day ##
To answer this question, we use again the functions of the *dplyr* library, grouping by *date* and summarising the data by the *steps* ignoring the *NA* values.

### Calculate of The Mean Steps per Day ###
```{r, echo= TRUE}
meanActbyDay <- actDS %>%
    group_by(date) %>%
    summarise(mean_steps= mean(steps, na.rm = TRUE))
```

To see the results in a convenient way, we can use the *xtable* library:
```{r, echo= TRUE, results= 'asis', warning= FALSE}
library(xtable, warn.conflicts= FALSE, quietly= TRUE)
meanActbyDay$date <- as.character(meanActbyDay$date)
actTable <- xtable(meanActbyDay)
print(actTable, type= "html")
```

### Histogram of the Total Number of Steps per Day ###
First, we need to do some calcutations again:
```{r, echo= TRUE}
totalActbyDay <- actDS %>%
    group_by(date) %>%
    summarise(total_steps= sum(steps, na.rm = TRUE))
```

We can do the histogram now:
```{r, echo= TRUE}
hist(totalActbyDay$total_steps, 
     breaks= 30, 
     col= "lightblue", 
     main = "Histogram of Total Steps by Day", 
     xlab= "Steps by Day")
```

### Report of the Mean and Median of the Total Number of Steps per Day ###
We can show the mean and median using *mean()* and *median()* functions but, instead of doing this way, we can achieve same result in a more simple way:
```{r, echo= TRUE}
summary(totalActbyDay$total_steps)
```

We can see that the mean of the total number of steps is `r format(round(mean(totalActbyDay$total_steps), 2), scientific= FALSE)` and the median is `r format(round(median(totalActbyDay$total_steps), 2), scientific= FALSE)`.

## Average Daily Activity Pattern ##
Again, we need to do some calcutations to know what's the average interval across all days:
```{r, echo= TRUE}
meanActbyInterval <- actDS %>%
    group_by(interval) %>%
    summarise(mean_steps= mean(steps, na.rm = TRUE))
```

### Time Series Plot ###
We'll use the *base* plotting system again, because this is a simple graphic:
```{r, echo= TRUE}
plot(x= meanActbyInterval$interval, 
     y= meanActbyInterval$mean_steps, 
     type= "l", 
     col= "blue", 
     main= "Time Series of Activity Pattern", 
     xlab= "5 minute interval", 
     ylab= "Average steps")
```

### Maximum Number of Steps ###
We can see that the `r meanActbyInterval$interval[which(meanActbyInterval$mean_steps == max(meanActbyInterval$mean_steps))]` interval is the highest one.

## Missing Values ##
There are an important number of missing values in the data. Until now, we ignoring it (for example, using *na.rm* attribute), but now we'll gonna see if they are introducing bias.

### Total Number of Missing Values ###
There are `r sum(is.na(actDS$steps))` missing values *(NA)* in the *steps* column of the data.

### Strategy to Avoid Missing Values ###
In order to avoid all the bias in the data introduced by the existing missing values, we choose to replace it by the mean activity by interval (previously calculated).

### Create a New Data Set with NA Filled In ###
```{r, echo= TRUE}
cleanActDS <- actDS %>%
    inner_join(meanActbyInterval, by= "interval")
cleanActDS$steps[is.na(cleanActDS$steps)] <- cleanActDS$mean_steps[is.na(cleanActDS$steps)]
```

### Histogram of the Total Number of Steps per Day (without NAs) ###
In order to do that, we need some calcutations again:
```{r, echo= TRUE}
totalNewActbyDay <- cleanActDS %>%
    group_by(date) %>%
    summarise(total_steps= sum(steps, na.rm = TRUE))
```

We can do the new histogram:
```{r, echo= TRUE}
hist(totalNewActbyDay$total_steps, 
     breaks= 30, 
     col= "lightgreen", 
     main = "Histogram of Total Steps by Day (without missing values)", 
     xlab= "Steps by Day")
```

### Mean and Median of the Total Number of Steps per Day Without Missing Values ###
We can see that the new mean of the total number of steps is `r format(round(mean(totalNewActbyDay$total_steps), 2), scientific= FALSE)` and the new median is `r format(round(median(totalNewActbyDay$total_steps), 2), scientific= FALSE)`.

As we expected, we noted that the new mean and median values are slightly different from the above ones, because now we have a largest population with more repeated central values.

## Differences in Activity Patterns Between Weekdays and Weekends ##
### New Factor Variable ###
I noticed that *weekdays()* function returns the day of the week name in the local language, so, in order to allow reproducibility, I opted for use the *is.weekend()* function, which is part of the *chron* package:

```{r, echo= TRUE}
library(chron, warn.conflicts= FALSE, quietly= TRUE)
weekActDS <- cleanActDS
weekActDS$weekend <- NA
weekActDS$weekend[which(is.weekend(weekActDS$date))] <- "weekend"
weekActDS$weekend[which(!is.weekend(weekActDS$date))] <- "weekday"
weekActDS$weekend <- factor(weekActDS$weekend)
table(weekActDS$weekend)
```

### Time Series Panel Plot ###
First, the calcutations:
```{r, echo= TRUE}
meanWeekendActbyInterval <- weekActDS %>%
    filter(weekend == "weekend") %>%
    group_by(interval) %>%
    summarise(mean_steps= mean(steps, na.rm = TRUE))

meanWeekdayActbyInterval <- weekActDS %>%
    filter(weekend == "weekday") %>%
    group_by(interval) %>%
    summarise(mean_steps= mean(steps, na.rm = TRUE))
```

Now, the plots:
```{r, echo= TRUE, fig.height= 8}
par(mfrow= c(2, 1))

plot(x= meanWeekendActbyInterval$interval, 
     y= meanWeekendActbyInterval$mean_steps, 
     type= "l", 
     col= "green", 
     main= "Activity Pattern (Weekend)", 
     xlab= "5 minute interval", 
     ylab= "Average steps")

plot(x= meanWeekdayActbyInterval$interval, 
     y= meanWeekdayActbyInterval$mean_steps, 
     type= "l", 
     col= "red", 
     main= "Activity Pattern (Weekday)", 
     xlab= "5 minute interval", 
     ylab= "Average steps")
```

*(Sorry for my english, I hope you understand me anyway ;-)*
